---
title: "Introduction to coding in R"
author: "Rebecca Barter"
format: html
embed-resources: true
---


## Using R as a calculator in the console

We can write R code directly in the console. 


### Exercise 

Use R as a calculator to compute `1 + 1` in the console







## Mathematical operations

You can do mathematical calculations in R using the following symbols:

* Parentheses: `(`, `)`
* Exponents: `^` or `**`
* Multiply: `*`
* Divide: `/`
* Add: `+`
* Subtract: `-`




### Exercise

- Compute 3 squared in the console

- Compute 2 divided by 100,000 in the console

- Use the logarithm function `log()` to compute the logarithm of 10 in the console










## Writing R code in a quarto document



The problem with only ever writing your R code in the console is that once you quit RStudio, there will be no record of the code that you ran.

Writing your code in a quarto help you to save your code and results in a *reproducible* way AND communicate your findings to other people.








### Code chunks


In a quarto document, like this one, you will write your r code in "code chunks" like this box below:

```{r}

```

For example, you can ask R to compute 1+1:

```{r}
1 + 1
```

You can view the "output" (result) of your code by either (a) *rendering* your quarto document, or (b) running your code in the *console*.


*Question:* Why is there a `[1]` before the output (`[1] 2`)? This is just specifying that `2` is the first "entry" of the output.



### Exercise

Create a new code chunk and compute 4 times 5

Inside your code chunk, on a new line, add some code to compute 3 squared (`3^2`). What do you think will happen when you render your document? Will both lines of code be computed, or just the final one?


```{r}
4 * 5
3^2
```

Both lines are computed in the rendered document!




## Code comments

R will ignore any text that follows a `#` symbol, so we can add "comments" to our code using `#` to make it easier to understand. 

```{r}
# compute 4 times 5
4 * 5
# compute 3 squared
3^2
```






## Chunk options

Take a look at the following code chunk. 

What do you think this code chunk will look like in the rendered html document?

``` {r}
#| echo: false
2 * 2
```









This `#|` syntax at the beginning of a code chunk corresponds to various options for when the code chunk is "rendered" into html (or pdf).

`#| echo: false` hides the code from the html output file, while still showing the output (`[1] 4`). 



### Exercise

Can you figure out what the output of the following code chunk will look like (will the code and code comment be shown in the output)? 

```{r}
#| echo: true
# two times three
2 * 3
```


The rendered document will show both the code and the output




