---
title: "Vectors"
author: "Rebecca Barter"
format: html
embed-resources: true
editor_options: 
  chunk_output_type: console
---


## Introduction to vectors

Previously, we were just defining simple objects such as 

```{r}
# define a variable x, containing 12
x <- 12
x
```


Let's create a vector that contains the ages of 5 people:


```{r}
# Use c() to define a variable age that contains the ages of 5 people at once
age <- c(12, 19, 22, 35, 18)
age
```

Let's ask what `age`'s type is:

```{r}
# identify the class of age
class(age)
```


Let's try and create a multi-vector consisting of multiple different types

```{r}
# try to create a vector, multi_vec, with numeric and character values
multi_vec <- c(1, 9, "banana", 10, -1)
multi_vec
```


Let's check the class of `multi_vec`

```{r}
# check the class of multi_vec
class(multi_vec)
```

This is an example of "type conversion".






Let's try to combine numeric and logical values in a vector

```{r}
# try to create a vector with numeric and logical values
multi_vec2 <- c(1, 5, TRUE, FALSE, -9)
multi_vec2
```

In type conversion, there is a hierarchy of types: **character > numeric > logical**.




### Exercise

What will be the output and type of the following code?

```{r}
vector_example <- c(TRUE, 4, "hello", FALSE, 0)
vector_example
# check the class
class(vector_example)
```








## Working with vectors

The following is an example of a "vectorized" operation:

```{r}
age <- c(12, 18, 22, 21, 17)
```


```{r}
# subtract 1 from age
age - 1
```

```{r}
# define a new vector, age2, corresponding to age * 2
age2 <- age * 2
age2
```


We can do computations with the two vectors together:

```{r}
# subtract age from age2
age2 - age
```



What happens when you try to do a computation with vectors of different lengths?

```{r}
# try to subtract the vector c(1, 2) from age
age - c(1, 2)
```






When you do mathematical computations on vectors of different lengths, the computation uses **recycling**

The above subtraction is equivalent to

```{r}
age - c(1, 2, 1, 2, 1)
```





### Vectorized logical operations


Let's ask which `age` entries are greater or equal to 18

```{r}
# ask which entries in age are greater than or equal to 18
age >= 18
```


Let's ask which entries are equal to 35

```{r}
# ask which entries in age are equal to 17
age == 17
```

Let's ask which entries are equal to 17 or 18

```{r}
# try to use == c(17, 18) to ask which entries are equal to 17 or 18
age == c(17, 18)
```

*This is only working because we got lucky with our recycling.*

This breaks if we switch the order of 18 and 17.

```{r}
# switch the order of 17 and 18
age == c(18, 17)
```

We can instead use the `%in%` operator:

```{r}
# use %in% to ask which entries in age are equal to 17 or 18
age %in% c(17, 18) 
```










## Functions for vectors

The `sum()` function can be used to add up all of the entries of a vector

```{r}
# compute the sum of all entries in age
sum(age)
```

The `mean()` function computes the mean:

```{r}
# compute the mean of all entries in age
mean(age)
```


The function `length()` tells us how many entries the vector contains

```{r}
# compute the length of the age vector
length(age)
```


We can use the `sum()` function to tally up the output of logical operations

```{r}
# compute the sum of the logical vector c(TRUE, FALSE, FALSE, TRUE)
sum(c(TRUE, FALSE, FALSE, TRUE))
```


```{r}
# add up the number of age entries that are 17 or 18
sum(age %in% c(17, 18))
```


```{r}
# add up the number of age entries that are over 18
sum(age > 18)
```






### Exercise

What is the *proportion* of entries in the `age` vector that are greater than 18?


```{r}

```


Two possible answers:

```{r}
sum(age > 18) / length(age)
```

```{r}
mean(age > 18)
```








## Indexing and subsetting vectors

We use the square bracket indexing syntax to extract entries from a vector

```{r}
# extract the first entry from age
age[1]
```

```{r}
# extract the fourth entry from age
age[4]
```

To extract the last entry in a vector:

```{r}
# extract the final entry from age
age[length(age)]
```






To remove an entry from a vector, you can use the `-` syntax.

```{r}
# remove the first entry from age
age[-1]
```

Has age changed?

```{r}
age
```

```{r}
# remove the fourth entry from age
age[-4]
```







### Extracting/removing multiple entries from a vector

Extract the first and third entries from `age`

```{r}
# extract entries 1 and 3 from age
age[c(1, 3)]
```

Remove the first and third entries

```{r}
# remove entries 1 and 3 from age
age[-c(1, 3)]
```







## Definining integer sequences

If we want to quickly define a sequence of integers, we can use the `:` syntax

```{r}
# use c() to create a vector 1, 2, 3, 4
c(1, 2, 3, 4)
# use the : syntax to create a vector 1, 2, 3, 4
1:4
# use the : syntax to create a vector of integers from 5 to 25
5:25
```

We can use these sequences to extract entries from vectors 

```{r}
# Use : to extract the first 4 entries from age
age[1:4]
```

What if I want to remove sequences of entries?

```{r}
# Use : to remove the first 4 entries from age
age[-(1:4)]
```






### The seq() function

The `seq()` function allows us to create a sequence of integers with non-unit increments

```{r}
# use seq() to create a sequence from 4 to 20 that increments by 2
# include argument names
seq(from = 4, to = 20, by = 2)
# exclude the argument names
seq(4, 20, 2)
```

```{r}
# use seq() to extract every second entry from age
age[seq(1, 5, 2)]
```






## Exercise


1. Create a vector that contains the values 5, 9, 2, 21, 34, 56, 2, -1, 5, 9

2. Extract the 4th and 9th entries

3. Extract the 1st and last entry

4. Extract every 3rd entry starting from the 2nd entry



```{r}
my_vec <- c(5, 9, 2, 21, 34, 56, 2, -1, 5, 9)
my_vec[c(4, 9)]
my_vec[c(1, length(my_vec))]
my_vec[seq(2, length(my_vec), 3)]
```









## Named vectors

You can provide names to the entries of a vector

```{r}
# Add the names "Dean", "Xiao", "Sara", "Ravi", "Maya" to the age vector
names(age) <- c("Dean", "Xiao", "Sara", "Ravi", "Maya")
age
```


You can extract/index entries from a vector using their names:

```{r}
# extract Ravi's age using named indexing
age["Ravi"]
```


```{r}
# extract Maya and Ravi's ages using named indexing
age[c("Maya", "Ravi")]
```










## Logical subsetting

We can use logical vectors to subset a vector


```{r}
# use a logical vector to extract the first, fourth, and fifth entries from age
age[c(TRUE, FALSE, FALSE, TRUE, TRUE)]
```





To identify which entries in a vector satisfy a criteria, I can ask a logical question. 

```{r}
# reminder: identify which entries in age are at least 18
age >= 18
```

and we can use logical expressions to subset a vector too

```{r}
# use age >= 18 to extract all ages that are at least 18
age[age >= 18]
```







### Subsetting with multiple conditions

How can we combine multiple conditions when subsetting, e.g., >= 17 and < 20

```{r}
#| error: true
# Try to combine the two conditions with a comma (,)
age[c(age >= 16, age < 20)]
```


We need to use the `&` operator

```{r}
# Identify which entries in age are both >= 16 and less than < 20
(age >= 16) & (age < 20)
```


```{r}
# What will be the output of the following & operations:
TRUE & TRUE
TRUE & FALSE
```

The "or" operator corresponds to the "|" 

```{r}
# What will be the output of the following | operations:
TRUE | TRUE
TRUE | FALSE
```


```{r}
# what do you think will happen when we ask if age <= 16 OR age > 20
age <= 16 | age > 20
```

```{r}
# extract all of the ages that are either <= 16 or > 20
age[age <= 16 | age > 20]
```



## Exercise

```{r}
vec <- c(4, 19, 2, 2, 3, 90, 55, 12)
```

Write some code for extracting the entries that are 

(a) less than 10

(b) less than 25 but greater than 10

(c) either less than 10 or equal to 55



```{r}
# a
vec[vec < 10]
# b
vec[(vec < 25) & (vec > 10)]
# c
vec[(vec < 10) | (vec == 55)]
```

