---
title: "Introduction to R objects"
author: "Rebecca Barter"
format: html
editor_options: 
  chunk_output_type: console
embed-resources: true
---

In this document, we will learn how to define "objects" in R.


## Defining objects


Let's assign the value 1 to an object called `x`

```{r}
# assign the value 1 to the variable x

```


We can view the value of `x` by writing it's name in a code chunk:


```{r}
# view x

```

Another way to assign a value to a variable is using "="

```{r}
# assign the value of 3 to the variable y using =

```


```{r}
# view y

```


You can also assign the output of mathematical calculations to a variable

```{r}
# assign the output of 1 + 1 to the variable z

# view z

```






### Exercise

As an exercise, your task is to define a new object `w` that contains the output of the product of 5 and 2. 

1. Create a new code chunk

2. Define the variable `w`, assign to it the output of the product of 5 and 2 

2. View the output of the object in the console.










## Working with variables


Let's do some computations with our variables


```{r}
# Define result to be the product of z and 7

```

```{r}
# update the value of z to be 4^2

```


**Question**: Do you think `result` will have changed?

Look at the values of `result` and `z`:

```{r}

```

```{r}

```










### Exercise

What will the output of the following code chunk be:


```{r}
value <- 1
computed_result <- value * 10 + 3^2
value <- value + 2
computed_result * 2 
```










## Variable names


Which of the following are valid R variable names?

```{r}
#| eval: false
min_height
max.height
_age
.mass
MaxLength
min-length
2widths
celsius2kelvin
```



The generally recommended variable name convention is: `variable_name`
